Associahedra for finite type cluster algebras \\ and minimal relations between $\b{g}$-vectors

### 2019-05-17 thirdPaper/draft/typeConeAssociahedra.tex
fixed missing part of graph associahedra

We have seen that $\{\tube, \tube'\}$ do not form an extremal pair unless both $\tube\ssm \tube'$ and $\tube'\ssm \tube$ are singletons. We now need to show that all these pairs are extremal.

Assume that~$\tube=\tube[s] \ssm \{v'\}$ and~$\tube'=\tube[s] \ssm \{v\}$ for some tube~$\tube[s] \in \tubes$ and distinct non-disconnecting vertices~$v,v'$ of~$\tube[s]$. To prove that this pair is extremal, we will construct a vector $\b{z}\in\R^{\tubes}$ such that \[\dotprod{\b{n}(\tube, \tube')}{\b{z}}=0,\]
but \[\dotprod{\b{n}(\tilde \tube, \tilde \tube')}{\b{z}}>0\] for any other pair $\{\tilde \tube, \tilde \tube'\}$ verifying the conditions. This will show that $\b{z}$ lies in the relative interior of the facet of the closed type cone defined by~$\{\tube, \tube'\}$, and hence that the inequality is not redundant.

\arnau{TODO! I have a partial construction, but there are still some undesired $\dotprod{\b{n}(\tilde \tube, \tilde \tube')}{\b{z}}=0$}

Define $\alpha(\tube,\tube')=\set{\tube[u]\in\tubes}{\tube[u]\subseteq \tube \cup \tube'\text{ and } v,v'\in \tube[u]}$, and set 
\[\b{z}_{\tube[s]}=-\big|\set{\tube[u]\in \tubes\ssm \alpha(\tube,\tube')\}}{\tube[u]\subseteq \tube[s]}\big|\]
for each tube~$\tube[s]\in\tubes$.

Notice that $\dotprod{\b{n}(\tube, \tube')}{\b{z}}=0$, and that 
\[\dotprod{\b{n}(\tilde \tube, \tilde \tube')}{\b{z}}= \big|\set{\tube[u]\in\tubes}{\tube[u]\subseteq\tube[t]\cup\tube[t']\text{ and }\tilde v,\tilde v'\in\tube[u]}\ssm \alpha(\tube,\tube')|\geq0,\]
where $\tilde v=\tilde \tube\ssm \tilde \tube'$  and $\tilde v'=\tilde \tube'\ssm \tilde \tube$. In particular, if $\tilde \tube\cup \tilde \tube'\notin\alpha(\tube,\tube')$ then $\dotprod{\b{n}(\tilde \tube, \tilde \tube')}{\b{z}}>0$.

\arnau{To finish. How to separate the remaining cases?}

@@

We derive from \cref{prop:extremalExchangeablePairsGA} and \cref{coro:extremalExchangeablePairsGA} all polytopal realizations of the nested fan, which we can explicitly express using one of the following two equivalent conditions.

@@

\begin{proof}
Note that any tube~$\tube$ with~$|\tube| \ge 2$ has two non-disconnecting vertices when it is a path, and at least three non-disconnecting vertices otherwise (the leaves of an arbitrary spanning tree of~$\tube$).
Therefore, each tube of~$\tubes$ which is not a singleton contributes to at least one extremal exchangeable pairs.
We conclude that the number of extremal exchangeable pairs is at least:
\[
|\tubes| - |V| = |\tubes \ssm \{\varnothing, \graphG\}| - (|V|-1) = N - n,
\]
with equality if and only if all tubes of~$\graphG$ are paths, \ie if and only if~$\graphG$ is a path.
\end{proof}

>>>>

We have seen that an exchangeable pair $\{\tube, \tube'\}$ is not extremal unless both $\tube\ssm \tube'$ and $\tube'\ssm \tube$ are singletons. We now need to show that all these pairs are extremal. Until we finish the proof, we will call such a pair a \defn{suspected extremal pair}.

Let $\{\tube, \tube'\}$ be a suspected extremal pair.
To prove that it is extremal, we will construct a vector $\b{w}\in\R^{\tubes}$ such that $\dotprod{\b{n}(\tube, \tube')}{\b{w}}< 0,$
but $\dotprod{\b{n}(\tilde \tube, \tilde \tube')}{\b{w}}>0$ for any other suspected extremal pair $\{\tilde \tube, \tilde \tube'\}$. This will show that the inequality induced by~$\{\tube, \tube'\}$ is not redundant.

Define $\alpha(\tube,\tube')=\set{\tube[u]\in\tubes}{\tube[u]\subseteq \tube \cup \tube'\text{ and } \tube\triangle\tube'\subseteq \tube[u]}$, where $\tube\triangle\tube'$ denotes the symmetric difference, and let $\b{x}\in\R^{\tubes}$ be such that
\[\b{x}_{\tube[s]}=-\big|\set{\tube[u]\in \tubes\ssm \alpha(\tube,\tube')}{\tube[u]\subseteq \tube[s]}\big|\]
for each tube~$\tube[s]\in\tubes$.

For any suspected extremal pair $\{\tilde \tube, \tilde \tube'\}$ we have
\[\dotprod{\b{n}(\tilde \tube, \tilde \tube')}{\b{x}}= \big|\alpha(\tilde \tube, \tilde \tube ')\ssm \alpha(\tube,\tube')|\geq0.\]
Notice that $\dotprod{\b{n}(\tilde \tube, \tilde \tube')}{\b{x}}=0$ if $\alpha(\tilde \tube, \tilde \tube ')\subseteq \alpha(\tube,\tube')$ and
$\dotprod{\b{n}(\tilde \tube, \tilde \tube')}{\b{x}}\geq 1$ otherwise.
In particular, if $\{\tilde \tube, \tilde \tube'\}\neq \{\tube, \tube'\}$ then
$\dotprod{\b{n}(\tilde \tube, \tilde \tube')}{\b{x}}\geq 1$ unless $\tilde \tube\cup \tilde \tube'\subsetneq \tube\cup \tube'$ because 
 $\tilde \tube\cup \tilde \tube'\in\alpha(\tilde \tube, \tilde \tube ')$, and if $\tilde \tube\cup \tilde \tube'=\tube\cup \tube'$ but $\{\tilde \tube, \tilde \tube'\}\neq \{\tube, \tube'\}$, then either $\tube$ or $\tube'$ will belong to $\alpha(\tilde \tube, \tilde \tube ')$.

Now, define $\b{y}\in\R^{\tubes}$ by \[\b{y}_{\tube[s]}=\begin{cases}
                             -1&\text{ if }|\tube[s]|\geq |\tube|\\
                             0&\text{ otherwise.}
                            \end{cases}\]
We have $\dotprod{\b{n}(\tube, \tube')}{\b{y}}=-1$. For any suspected extremal pair $\{\tilde \tube, \tilde \tube'\}$ we have $\dotprod{\b{n}(\tube, \tube')}{\b{y}}\geq -1$. If moreover $\tilde \tube\cup \tilde \tube'\subsetneq \tube\cup \tube'$, then we have $\dotprod{\b{n}(\tube, \tube')}{\b{y}}\geq 0$.

Finally, consider $\b{z}\in\R^{\tubes}$ given by
\[\b{z}_{\tube[s]}=-\big|\set{\tube[u]\in \alpha(\tube,\tube')}{\tube[u]\subseteq \tube[s]}\big|. \]
It verifies $\dotprod{\b{n}(\tilde \tube, \tilde \tube')}{\b{z}}\geq0$ for any suspected extremal pair~$\{\tilde \tube, \tilde \tube'\}$, and $\dotprod{\b{n}(\tube, \tube')}{\b{z}}=|\alpha(\tilde \tube, \tilde \tube ')|>0$ if $\alpha(\tilde \tube, \tilde \tube ')\subseteq \alpha(\tube,\tube')$.

For $0<\varepsilon<1$, and $0<\delta<\frac{\varepsilon}{|\alpha(\tube,\tube')|}$, the vector $\b{w}=\b{x}+\varepsilon \b{y}+\delta \b{z}$ fulfills the desired properties.
Indeed, if $\{\tilde \tube, \tilde \tube'\}\neq \{\tube,\tube '\}$ is a suspected extremal pair with $\alpha(\tilde \tube, \tilde \tube ')\not \subseteq \alpha(\tube,\tube')$, then 
\[\dotprod{\b{n}(\tilde \tube, \tilde \tube')}{\b{w}}=\underbrace{\dotprod{\b{n}(\tilde \tube, \tilde \tube')}{\b{x}}}_{\geq 1}+\underbrace{\varepsilon \dotprod{\b{n}(\tilde \tube, \tilde \tube')}{\b{y}}}_{> -1}+\underbrace{\delta \dotprod{\b{n}(\tilde \tube, \tilde \tube')}{\b{z}}}_{\geq 0}> 0.\]
 If $\{\tilde \tube, \tilde \tube'\}\neq \{\tube,\tube '\}$ is a suspected extremal pair with $\alpha(\tilde \tube, \tilde \tube ')\subseteq \alpha(\tube,\tube')$, then $\tilde \tube\cup \tilde \tube'\subsetneq \tube\cup \tube'$ and 
\[\dotprod{\b{n}(\tilde \tube, \tilde \tube')}{\b{w}}=\underbrace{\dotprod{\b{n}(\tilde \tube, \tilde \tube')}{\b{x}}}_{=0}+\underbrace{\varepsilon \dotprod{\b{n}(\tilde \tube, \tilde \tube')}{\b{y}}}_{\geq 0}+\underbrace{\delta \dotprod{\b{n}(\tilde \tube, \tilde \tube')}{\b{z}}}_{>0}> 0.\]
However,
\[\dotprod{\b{n}(\tube, \tube')}{\b{w}}=\underbrace{\dotprod{\b{n}(\tube, \tube')}{\b{x}}}_{=0}+\underbrace{\varepsilon \dotprod{\b{n}(\tube, \tube')}{\b{y}}}_{=-\varepsilon}+\underbrace{\delta \dotprod{\b{n}( \tube,  \tube')}{\b{z}}}_{=\delta|\alpha(\tube ,\tube')|<\varepsilon}< 0.\]

@@

We derive from \cref{prop:extremalExchangeablePairsGA} and \cref{coro:extremalExchangeablePairsGA} all polytopal realizations of the nested fan, which we can explicitly express.
% using one of the following two equivalent conditions.

@@

\begin{proof}
Note that any tube~$\tube$ with~$|\tube| \ge 2$ has two non-disconnecting vertices when it is a path, and at least three non-disconnecting vertices otherwise (the leaves of an arbitrary spanning tree of~$\tube$, or any vertex if it is a cycle).
Therefore, each tube of~$\tubes$ which is not a singleton contributes to at least one extremal exchangeable pair.
We conclude that the number of extremal exchangeable pairs is at least:
\[
|\tubes| - |V| = |\tubes \ssm \{\varnothing, \graphG\}| - (|V|-1) = N - n,
\]
with equality if and only if all tubes of~$\graphG$ are paths, \ie if and only if~$\graphG$ is a path.
\end{proof}
